---
title: "Analysis of Tooth Growth"
author: "Frank Jung"
graphics: yes
output:
  pdf_document:
    fig_crop: no
    fig_height: 4.5
    latex_engine: xelatex
    number_sections: yes
    toc: yes
  html_document:
    toc: yes
classoption: legalpaper
---

# Synopsis

In this project we will analyse the 
[ToothGrowth](https://stat.ethz.ch/R-manual/R-devel/library/datasets/html/ToothGrowth.html)
dataset to compare odontoblast growth by supplement and dosage. The question we 
want to answer is: Does the delivery method of supplements effect mean 
[odontoblast](https://en.wikipedia.org/wiki/Odontoblast) growth? i.e. Is there 
any measurable difference in means of growth between these supplements and
dosages?
```{r setoptions, echo=FALSE, message=FALSE, warning=FALSE}
require(datasets)
require(knitr, quietly = TRUE)
require(ggplot2, quietly = TRUE)
knitr::opts_chunk$set(echo = FALSE, cache = TRUE, cache.path = "cache/", fig.path = "figure/", fig.align="center")
```

All code for this project is available on [GitHub](https://github.com) from [here](https://github.com/frankhjung/statinference-project2).

# Exploratory Analysis

First, let us explore the data:
```{r exploredataframe, echo=TRUE, message=FALSE, warning=FALSE}
# for convenience save ToothGrowth into data frame tg
tg <- data.frame(ToothGrowth)
tg$dose <- factor(tg$dose)
str(tg)
table(tg[tg$supp == "OJ", c("dose")])
table(tg[tg$supp == "VC", c("dose")])
cat(range(tg$len))
```

Below is a violin plot by supplement and dose:
```{r exploreplot, echo=FALSE, message=FALSE, warning=FALSE}
levels(tg$supp)[levels(tg$supp) == "OJ"] <- "Orange Juice (OJ)"
levels(tg$supp)[levels(tg$supp) == "VC"] <- "Vitamin C (VC)"
ggplot(tg, aes(x = dose, y = len)) +
    geom_violin() +
    geom_boxplot(width = 0.2, fill = "purple", alpha = 0.4) +
    stat_summary(fun.y = mean, geom = "point", colour = "white", shape = 19, size = 2) +
    facet_grid(supp ~ .) +
    labs(x = "Dose (milligrams)", y = "Odontoblast Length (microns)") +
    ggtitle("Kernel Probability Density of Length by Dosage")
```

In summary, the ToothGrowth dataset is small, with:

* 10 observations of odontoblast length (microns) for each dose
* 3 distinct dosage levels (milligrams)
* 2 supplement types (VC = Vitamin C and OJ = Orange Juice)

The data suggests that there may be observable differences with lower dosages.

# Assumptions

In this analysis we assume that:

* the observations are independent ([different groups of guinea-pigs](http://jn.nutrition.org/content/33/5/491.full.pdf))
* the data is assumed to be normally distributed ([likely, as this is a biological measurement](https://en.wikipedia.org/wiki/Normal_distribution#/Approximate_normality))

# Test Hypothesis

Our null hypothesis is that there is no difference in odontoblast mean lengths 
between supplement types for each dosage. That is: $H_0 : \mu_{oj} = \mu_{vc}$ 
with an alternative hypothesis: $H_a : \mu_{oj} \ne \mu_{vc}$. To test this we 
will calculate a 95% confidence interval, using a two-sided T test of
independent groups, with variances assumed to be unequal.

For each dose, compare mean lengths from each supplement. Report the 95%
confidence interval and p-value:
```{r testhypothesis, echo=TRUE, message=FALSE, warning=FALSE}
by(tg, tg$dose,
   function (x) t.test(len ~ supp, x, paired = F, var.equal = F, alternative = "t"))
```

Note that the confidence interval for supplement dosages of 0.5 and 1.0 
milligrams does not contain 0. In addition, the p-values for these observations 
are much less than 5%. Compare this to the confidence interval for supplement
dosages of 2.0 milligrams, which does contain 0 and the p-value is above 96%.

# Conclusion

We are testing the null hypothesis that odontoblast mean lengths are the same, 
regardless of the supplement type or dosage: $H_0 : \mu_{oj} = \mu_{vc}$. The
alternative hypothesis is that they are not equal: $H_a : \mu_{oj} \ne
\mu_{vc}$.

From the [p values](https://en.wikipedia.org/wiki/P-value) and [confidence
intervals](https://en.wikipedia.org/wiki/Confidence_interval) we can conclude:

| Dosage (milligrams) | Comparison of mean lengths between supplements |
|:-------------------:|------------------------------------------------|
| 0.5                 | We reject the null hypothesis                  |
| 1.0                 | We reject the null hypothesis                  |
| 2.0                 | We fail to reject the null hypothesis          |

So, it appears that effects on odontoblast mean length diminishes between
supplement types at higher dosages.

# References

* [Crampton, E. W. (1947) The growth of the odontoblast of the incisor teeth as a criterion of vitamin C intake of the guinea pig. The Journal of Nutrition.](http://jn.nutrition.org/content/33/5/491.full.pdf)
* [Normal Distribution. Wikipedia](https://en.wikipedia.org/wiki/Normal_distribution)
* [Odontoblast. Wikipedia](https://en.wikipedia.org/wiki/Odontoblast)
* [ToothGrowth dataset](https://stat.ethz.ch/R-manual/R-devel/library/datasets/html/ToothGrowth.html)
* [GitHub repository containing project code](https://github.com/frankhjung/statinference-project2)

$\hfill \square$
